Detection and characterisation of stellar multiplicity with Gaia

### 2018-06-29 article/ms.tex
minor

\newcommand{\NumberOfHigherOrderSystems}{XXX}
\newcommand{\NumberOfSingleLinedSystems}{XXX}

@@

\title{Detection and characterisation of stellar multiplicity with \Gaia}

@@

@@

\begin{abstract}
Most stars have stellar companions. The occurrence rate and properties of stellar 
multiples (e.g., binaries, triples) is relevant to many fields of astrophysics, 
but the effects of stellar multiplicity are commonly ignored in the analysis of 
most stellar and galactic surveys. Here we use \Gaia\ data to identify 
\todo{\NumberOfHigherOrderSystems} stellar multiple systems from radial velocity
and astrometric jitter, photometry, and intrinsic luminosity. Our sample includes
\todo{\NumberOfSingleLinedSystems} unresolved binary systems detected solely
from radial velocity or astrometric jitter, which have orbital periods up to 
$\approx$3.5\,yrs. We show that the ratio of jitter in radial velocity and
astrometry can be calibrated to estimate inclination angles (and companion masses)
for thousands of systems. \todo{We comment on stellar multiplicity with stellar
metallicity}. \todo{We comment on systems with stellar remnants (e.g. NS/BHs).}
\end{abstract}

\keywords{stars: dynamics}

@@

The \Gaia\ space telescope provides exquisite astrometry, photometry, and radial
velocity measurements over many years for million of point stars in our galaxy.
%These data provides an excellent opportunity to self-calibrate the errors in those
%measurements. 
When a point source has an unresolved stellar companion within some range of orbital
and stellar parameters, it is expected that there will be an anomalous excess in 
these measured properties relative to a single star of a similar type. This is
nothing new: astronomers have used radial velocity \citep{firstplanet} and 
astrometric variations \citep{someone} and photometry \citep{someone} to infer the 
presence of stellar and exoplanet companions forever. However, the opportunity
presented by \Gaia\ is to infer the presence of stellar companions for millions
of point sources, across both hemispheres. No ground- or space-based instrument
has ever provided such an opportunity.

@@

\section{Method}

@@

Radial velocity measurements are not available for individual epochs in the 
second \Gaia\ data release \citep{Lindegren:2018,Gaia_RVS}, but the a radial velocity point estimate and error
is available for 7,224,631 sources \cite{gaia_rvs}. 
The radial velocity error provided by \Gaia\ ($\GaiaRVE$; column name \texttt{radial\_velocity\_error}) 
is a function of the number of transits $N$ (i.e., the number of observations; 
represented by column \texttt{rv\_nb\_transits}) and the scatter among those $N$ measurements. 
This allows us to calculate the scatter among those measurements, which we will
refer to as the \emph{radial velocity jitter}, for each source
\begin{eqnarray}
\RVJitter = \sqrt{\frac{2N}{\pi}}\GaiaRVE \quad .
\end{eqnarray}

@@

Many higher-order star systems are unresolved point sources in \Gaia, but the
stellar companion(s) have sufficient mass to measurably affect the tangential 
velocity of the primary and produce a jitter in astrometric positions. There are
many other sources of astrometric jitter, but here we show that the principle
effect of astrometric excess noise is due to stellar companions. Note that due
to the astrometry `degree of freedom bug' in the processing of the second \Gaia\
data release \citep[e.g., see Appendix A1 of ][]{Lindegren:2018}, we only make
use of astrometric excess noise from the \todo{Tycho-Gaia Astrometric Solution}
\citep{Michalik, TGAS, etc}.

@@

\begin{figure*}
	\includegraphics[width=1.0\textwidth]{../figures/sb9-comparison.pdf}
    \caption{Source radial velocity variance shown with respect to flux in
    the \Gaia\ BP, G, and RP bands. The source radial velocity variance 
    most closely follows the RP flux as it is closer to the \Gaia\ RVS
    band.}
\end{figure*}

\begin{figure}
	\includegraphics[width=0.5\textwidth]{../figures/sb9-comparison.pdf}
    \caption{Source mean \Gaia\ RP flux and radial velocity variance.
    Points are represented by their mean $BP-RP$ color, confirming that
    there is some color dependency on the source radial velocity variance.}
    \label{fig:rp-flux-color}
\end{figure}

\begin{figure}
	\includegraphics[width=0.5\textwidth]{../figures/sb9-comparison.pdf}
    \caption{Mixture model fitted.}
    \label{fig:mixture-model}
\end{figure}

\begin{figure}
	\includegraphics[width=0.5\textwidth]{../figures/sb9-comparison.pdf}
    \caption{Detected binary fraction as a function of the mean RP flux
    (top panels), the $BP-RP$ color (second panels), the number of 
    radial velocity transits (third panels), and the inferred temperature
    (bottom panels). No trend is observed with mean RP flux or the number 
    of radial velocity transits. However, our model finds bluer (hotter)
    stars are more likely to be binaries, despite including color as a
    dependency in our model.}
    \label{fig:binary-fraction-wrt-params}
\end{figure}

@@

\subsection{Comparison with \citet{El_Badry:2017,El_Badry:2018}}

\subsection{Comparison with \citet{Badenes:2017}}

@@

\subsection{Comparison with \citet{Troup:2010}}

\subsection{Comparison with \citet{Price-Whelan:2010}}

@@

It is a pleasure to thank
    Adrian Price-Whelan (Princeton University),
	Lorenzo Spina (Monash University), 
		and 
	Kevin C. Schlaufman (Johns Hopkins University).
This work was partly supported through the Australian Research Council 
through Discovery Grant DP160100637.
This work has made use of data from the European Space Agency (ESA) mission {\it
Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia} Data
Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
has been provided by national institutions, in particular the institutions
participating in the {\it Gaia} Multilateral Agreement.  This research was
started at the NYC Gaia DR2 Workshop at the Center for Computational
Astrophysics of the Flatiron Institute in 2018 April.

@@

\software{
	\package{Astropy} \citep{astropy},
    \package{IPython} \citep{ipython},
    \package{matplotlib} \citep{mpl},
    \package{numpy} \citep{numpy},
    \package{scipy} \citep{scipy},
    \package{Stan} \citep{stan},
    \package{CosmoHub} \citep{cosmohub}
    %tensorflow
}

>>>>

\newcommand{\NumberOfStellarMultiples}{XXX}
\newcommand{\NumberOfStellarSingles}{XXX}

@@

\title{Detection and characterisation of stellar multiplicity with Gaia}

@@

% Some order of the following authors (order is TBD)

@@

\author[0000-0003-3494-343X]{Carles Badenes}
\affiliation{Department of Physics and Astronomy, 
			 University of Pittsburgh, 
			 3941 O'Hara Street, Pittsburgh, PA 15260, USA}

\author[0000-0003-0872-7098]{Adrian M. Price-Whelan}
\affiliation{Department of Astrophysical Sciences, 
			 Princeton University, 
			 Princeton, NJ 08544, USA}

\begin{abstract}
The properties and occurrence rate of stellar multiples (e.g., binaries, triples) 
underpins much of astrophysics. Precisely measuring these quantities is challenging
and observationally demanding. Here we use \Gaia\ data to detect and partially 
characterise \NumberOfStellarMultiples\ stellar multiple systems based on excess 
astrometric jitter, excess radial velocity jitter, the absence of reported radial
velocities, photometry, and intrinsic luminosity.
\todo{We show that astrometric and radial velocity jitter can be calibrated to
estimate inclination angles (and directly constrain companion masses) for thousands 
of systems.}
Comparisons with literature surveys on stellar multiplicity demonstrate that we 
reliably detect binary systems with orbital periods up to $\approx$3.5\,yr. 
\todo{A comment on stellar multiplicity with stellar metallicity.}
\todo{A comment on stellar multiplicity in the field relative to clusters.}
\todo{A comment on systems with stellar remnants (e.g., NS/BHs).}
\end{abstract}

\keywords{(stars:) binaries: general, (stars:) binaries: spectroscopic, (stars:) binaries (including multiple): close, astrometry, techniques: radial velocities, methods: statistical}

@@

The \Gaia\ space telescope provides exquisite astrometry, photometry, and radial
velocity measurements over many years for million of point stars in our galaxy.
%These data provides an excellent opportunity to self-calibrate the errors in those
%measurements. 
When a point source has an unresolved stellar companion within some range of orbital
and stellar parameters, it is expected that there will be an anomalous excess in 
these measured properties relative to a single star of a similar type. This is
nothing new: astronomers have used radial velocity \citep{Butler:1996} and 
astrometric variations \citep{Muterspaugh:2010} and photometry \citep{someone} to infer the 
presence of stellar and exoplanet companions forever. However, the opportunity
presented by \Gaia\ is to infer the presence of stellar companions for millions
of point sources, across both hemispheres. No ground- or space-based instrument
has ever provided such an opportunity.

@@

\section{Method} \label{sec:method}

@@

Radial velocity measurements are not available for individual epochs in the 
second \Gaia\ data release \citep{Lindegren:2018,Cropper:2018}, but the a radial velocity point estimate and error
is available for 7,224,631 sources \cite{Cropper:2018}. 
The radial velocity error provided by \Gaia\ ($\GaiaRVE$; column name \texttt{radial\_velocity\_error}) 
is a function of the number of transits $N$ (i.e., the number of observations; 
represented by column \texttt{rv\_nb\_transits}) and the scatter among those $N$ measurements. 
This allows us to calculate the scatter among those measurements, which we will
refer to as the \emph{radial velocity jitter}, for each source
\begin{eqnarray}
\RVJitter = \sqrt{\frac{2N}{\pi}}\GaiaRVE \quad .
\end{eqnarray}

@@

Many higher-order star systems are unresolved point sources in \Gaia, but the
stellar companion(s) have sufficient mass to measurably affect the tangential 
velocity of the primary and produce a jitter in astrometric positions. There are
many other sources of astrometric jitter, but here we show that the principle
effect of astrometric excess noise is due to stellar companions. Note that due
to the astrometry `degree of freedom bug' in the processing of the second \Gaia\
data release \citep[e.g., see Appendix A1 of ][]{Lindegren:2018}, we only make
use of astrometric excess noise from the \todo{Tycho-Gaia Astrometric Solution}
\citep{Michalik:2015, Lindegren:2016}.

@@

\begin{figure*}
	\includegraphics[width=1.0\textwidth]{../figures/todo.png}
    \caption{Source radial velocity variance shown with respect to flux in
    the \Gaia\ BP, G, and RP bands. The source radial velocity variance 
    most closely follows the RP flux as it is closer to the \Gaia\ RVS
    band.}
\end{figure*}

\begin{figure}
	\includegraphics[width=0.5\textwidth]{../figures/todo.png}
    \caption{Source mean \Gaia\ RP flux and radial velocity variance.
    Points are represented by their mean $BP-RP$ color, confirming that
    there is some color dependency on the source radial velocity variance.}
    \label{fig:rp-flux-color}
\end{figure}

\begin{figure}
	\includegraphics[width=0.5\textwidth]{../figures/todo.png}
    \caption{Mixture model fitted.}
    \label{fig:mixture-model}
\end{figure}

\begin{figure}
	\includegraphics[width=0.5\textwidth]{../figures/todo.png}
    \caption{Detected binary fraction as a function of the mean RP flux
    (top panels), the $BP-RP$ color (second panels), the number of 
    radial velocity transits (third panels), and the inferred temperature
    (bottom panels). No trend is observed with mean RP flux or the number 
    of radial velocity transits. However, our model finds bluer (hotter)
    stars are more likely to be binaries, despite including color as a
    dependency in our model.}
    \label{fig:binary-fraction-wrt-params}
\end{figure}

@@

\subsection{Comparison with \citet{El-Badry:2018}}

\subsection{Comparison with \citet{Badenes:2018}}

@@

\subsection{Comparison with \citet{Troup:2016}}

\subsection{Comparison with \citet{Price-Whelan:2018}}

@@

% At least some of these people will be promoted to the author list.
It is a pleasure to thank
	Berry Holl (Geneva),
	Jose Hernandez (ESAC),
	Daniel Michalik (ESA/ESTEC),
	Kevin C. Schlaufman (Johns Hopkins University),
		and
	Lorenzo Spina (Monash University).
This work was supported in part by the Australian Research Council 
through Discovery Grant DP160100637.
This work has made use of data from the European Space Agency (ESA) mission {\it
Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia} Data
Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
has been provided by national institutions, in particular the institutions
participating in the {\it Gaia} Multilateral Agreement.  This research was
developed in part at the NYC Gaia DR2 Workshop at the Center for Computational
Astrophysics of the Flatiron Institute in 2018 April.

@@

\software{
	\package{Astropy} \citep{astropy},
    \package{IPython} \citep{ipython},
    \package{matplotlib} \citep{mpl},
    \package{numpy} \citep{numpy},
    \package{scipy} \citep{scipy},
    \package{Stan} \citep{stan},
    \package{CosmoHub} \citep{cosmohub},
    \package{TensorFlow} \citep{tensorflow}
    \package{Jupyter Notebooks} \citep{jupyter-notebooks}
}
